---
title: Untitled
subtitle: A slightly longer title
format:
  360-analysis-html: default
author: James Goldie
date: last-modified
code-fold: true
---

## Introduction

```{r}
#| label: setup

library(tidyverse)
library(sf)
library(terra)
library(here)
```

If they aren't already present, we'll download the Bedmap2 topography/bathymetry grids.

```{r}
#| label: download-bedmap2
download_if_not_present <- function(url, dest_zip, dest_dir,
  temp_timeout = 3600) {
  if (!file.exists(dest_zip)) {
    message(paste0("Downloading data: ", basename(url), "..."))
    old_timeout <- options("timeout")
    options(timeout = temp_timeout)
    download.file(url, dest_zip)
    unzip(dest_zip, exdir = dest_dir)
    options(timeout = old_timeout)
  } 
}

# bedmap2:
# - https://tc.copernicus.org/articles/7/375/2013/tc-7-375-2013.html
# - https://secure.antarctica.ac.uk/data/bedmap2/
# - readme: https://secure.antarctica.ac.uk/data/bedmap2/bedmap2_readme.txt
download_if_not_present(
  "https://secure.antarctica.ac.uk/data/bedmap2/bedmap2_tiff.zip",
  here("data", "bedmap2.zip"),
  here("data", "bedmap2"))
```

We want to load these GeoTIFFs in to the browser, but they're way too big for users to download. Let's resample them to make smaller files:

```{r}
#| label: process-tiffs

# load a GeoTIFF, shrink it and write the smaller version out
shrink_raster <- function(x) {
  here("data", "bedmap2", "bedmap2_tiff", paste0("bedmap2_", x, ".tif")) |>
    rast() |>
    aggregate(fact = 12, fun = "mean") |>
    writeRaster(
      here("data", "bedmap2", paste0("bedmap2_", x, "_small.tif")),
      overwrite = TRUE)
}

get_outlines <- function(x, height) {
  here("data", "bedmap2", "bedmap2_tiff", paste0("bedmap2_", x, ".tif")) |>
    rast() |>
    (\(x)(x >= height))() |>
    as.polygons() |>
    subset(TRUE) |>
    project("epsg:4326") |>
    writeVector(
      here("data", "bedmap2", paste0("bedmap2_", x, "_outline.geojson")),
      filetype = "GeoJSON",
      overwrite = TRUE)
}

# run the above operations on our tiffs
c("surface", "bed", "thickness") |> walk(shrink_raster)
c("surface", "bed") |> walk2(c(1, 0), get_outlines)
```

:::{.callout-note}
Note that our GeoJSON outlines are _right-handed_, which most tools expect. But D3.js [requires _left-handed polygons](https://gis.stackexchange.com/questions/392452/why-d3-js-works-only-with-geojson-violating-right-hand-rule), so we'll need to use turf.js to fix them in the browser.
:::
